---
title: "Exercise Solutions: Joining and Reshaping"
author:
- name: University of North Carolina at Chapel Hill
output: 
  html_document:
    theme: spacelab
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

**Download Data** <br>
[Download data](data/brazilian-ecommerce-subset.zip)
<br><br>

* Extract the data within the data directory.  You should end up with a folder called "brazilian-ecommerce".
```{r download, message=FALSE}
# Load the tidyverse package.
library(tidyverse) 
```


### 1. Read olist_public_dataset_v2.csv into R. Explore the dataset. If necessary, refer to the metadata provided [here](https://www.kaggle.com/olistbr/brazilian-ecommerce).

**Answer:**  
```{r}
data <- read_csv("data/brazilian-ecommerce/olist_public_dataset_v2.csv")
summary(data)
```


### 2. Read product_category_name_translation.csv into R.  Merge this dataframe into olist_public_dataset_v2.csv using `product_category_name` as the key variable.

**Answer:**  
```{r}
trans <- read_csv("data/brazilian-ecommerce/product_category_name_translation.csv")

data_tr <- left_join(data, trans, by=c("product_category_name"="product_category_name"))
```


Let's explore which products are most frequently purchased together:

### 3. Use `group_by` and `summarize` to find the total `order_items_qty` for each `product_category_name_english` in each `customer state`.

[Review](https://unc-libraries-data.github.io/R-Open-Labs/week3_Transformations/R_OpenLabs_3_Transformations.html)

**Answer:**  
```{r}
products_long <- data_tr %>% group_by(product_category_name_english, customer_state) %>% 
  summarize(total_orders = sum(order_items_qty))
```


### 4. Use `pivot_wider` to create a new dataframe with a row for each `customer_state` and a column for each `product_category_name_english`.  Name this dataframe `products`.

**Answer:**  
```{r}
products <- pivot_wider(products_long, id=customer_state, names_from=product_category_name_english, values_from=total_orders)
```


### 5. Run the two lines of code below (make sure your dataframe from step 4 is called `products`!)

```
products <- ungroup(products) #remove grouping 
products[is.na(products)] <- 0 #replace missing data with zeroes
```
**Answer:**  
```{r}
products <- ungroup(products) #remove grouping 
products[is.na(products)] <- 0 #replace missing data with zeroes
```


### 6. Use `ggpairs` or other Exploratory Data Analysis techniques to look for relationships between purchases of `small_appliances`,`consoles_games`,`air_conditioning`, and `construction_tools_safety`. (Remember to run `library(GGally)` before using `ggpairs`).

**Answer:**  
```{r fig.width=6, fig.height=6}
library(GGally)
products %>% select(small_appliances, consoles_games, air_conditioning, construction_tools_safety) %>% 
  ggpairs()
```

Unsurpsisingly, there is a positive relationship between each pair of variables.  This probably represents the underlying influence of popluation - states with larger populations probably order more of everything!

### 7. Repeat problems 3-6 with `order_products_value` (i.e. the amount spent vs the quantity purchased).  Do you see different patterns?  Explore other product categories.

**Answer:**  
```{r}
prod_value <- data_tr %>% group_by(product_category_name_english, customer_state) %>% 
  summarize(total_value = sum(order_products_value)) %>% 
  pivot_wider(id=customer_state, names_from=product_category_name_english, values_from=total_value) %>% 
  ungroup()

prod_value[is.na(prod_value)] <- 0

prod_value %>% select(small_appliances, consoles_games, air_conditioning, construction_tools_safety) %>% 
  ggpairs()

```
We see similar, if a bit weaker, relationships.

### 8. Use `pivot_longer` and then `pivot_wider` to convert your `products` dataframe to one where each row is a different `product_category_name_english` and each column represents a different `customer_state`.  

**Answer:**  
```{r}
prod_by_state <- products %>% pivot_longer(cols=-customer_state) %>% 
  pivot_wider(names_from = customer_state, values_from =value)
```


### 9. Choose 5 states. Which of these states have the most similar patterns of spending (as measured by correlation)?

**Answer:**  
```{r}
chosen_states <- c("RN", "SC", "DF", "PE", "RR")
prod_by_state %>% select(chosen_states) %>% ggpairs()
```

Among these states, the most similar pair is DF and SC with a correlation of 0.963.