---
title: "Exercise Solutions: Data Transformations"
author:
- name: University of North Carolina at Chapel Hill
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
    theme: spacelab
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

**Download Data** <br>
[Five Thousand Wine Reviews](https://unc-libraries-data.github.io/R-Open-Labs/week3_Transformations/data/5k_wine_reviews.csv)
<br><br>

```{r download, message=FALSE}
# Load the tidyverse package.
library(tidyverse) 

# Read in the data and store it as a data frame object called "review".
reviews <- read_csv("data/5k_wine_reviews.csv")

```

## 1. What is the price and review score for the most expensive French wine?

**Answer**<br>
The most expensive French wine in this data set is the 2009 Château Margaux, which costs $1900 and has a score of 98 points.

```{r one, message=FALSE}

# slice() is a dplyr convenience function that extracts a particular row
reviews %>%
  filter(country == "France") %>%
  arrange(desc(price)) %>%
  select(title, price, points) %>%
  slice(1)

```


## 2. How many countries have wine that cost more than $500?

**Answer**<br>

There are two countries that have wine that costs more than $500.

```{r two, message=FALSE}

# If we want the names of the countries, we can use unique(), a base R function
reviews %>%
  filter(price  > 500) %>%
  select(country) %>%
  unique()

# If we just want the count, we can use n_distinct(), a dplyr helper function that
# is a shortcut for length(unique(x))
reviews %>%
  filter(price  > 500) %>%
  select(country) %>%
  n_distinct()



```



## 3. To which country and province should you travel to find the cheapest Tempranillo?

**Answer**<br>

The cheapest Temperanillo can be found in the Mendoza Province of Argentina.

```{r three, message=FALSE}

# If we want the names of the countries, we can use unique(), a base R function
reviews %>%
  filter(variety == "Tempranillo") %>%
  arrange(price) %>%
  select(country, province, price) %>%
  slice(1)

```


## 4. You found a website that lets you order any wine for a flat shipping fee of $15! Find the total price for each wine that includes shipping.

**Answer**<br>
There are many ways to do this (as always in R), here's one way to accomplish this using the mutate() function.


```{r four, message=FALSE}

reviews <- reviews %>%
  mutate(total_price = price + 15)

```


## 5. The following code produces errors. Correct them all:

```{r errors, eval=FALSE}

reviews %>%
  # Title should be lowercase to match column name in dataframe
  # The == is for testing logical equality, should not go between function name and parentheses
  select == (Title, province, country, price) %>% 
  # The = is for assignment. This should be == to see if country is Germany
  # Germany should be in quotes since it is a character string
  # The pipe symbol is missing the closing %
  filter (country = Germany) %> 
  group_by (province) %>%
  # Summarise should be lower case
  Summarise (mean_price == mean(price))

```

**Answer**<br>
In addition to the errors pointed out below, there's also a [coding style](https://style.tidyverse.org/syntax.html#spacing) difference here with respect to the spacing between function names and `()`. The spaces after function names isn't necessarily wrong, but it is often frowned upon as a matter of style.


```{r, message=FALSE}
reviews %>%
  select(title, province, country, price) %>%
  filter(country == "Germany") %>%
  group_by(province) %>%
  summarise(mean_price = mean(price))

```