---
title: "Exercise Solutions: EDA"
author:
- name: University of North Carolina at Chapel Hill
output: 
  html_document:
    theme: spacelab
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

**Download Data** <br>
[Boston AirBnB Data](data/Boston_listings.csv)
<br><br>

```{r download, message=FALSE}
# Load the tidyverse package.
library(tidyverse) 

# Read in the data and store it as a data frame object called "listings".
listings <- read_csv("data/Boston_listings.csv")

```


### 1. Make a scatterplot of review_scores_rating vs reviews_per_month.  Add a best fit linear regression line with `geom_smooth` (check the help with `?geom_smooth`, or refer back to [week one](https://unc-libraries-data.github.io/R-Open-Labs/week1_Introduction/R_OpenLabs_1_Introduction.html))

**Answer**

Just the scatterplot:
```{r fig.width=5, fig.height=3}
ggplot(listings, aes(x=review_scores_rating, y=reviews_per_month)) + 
  geom_point()
```

Scatterplot with linear regression line:
```{r fig.width=5, fig.height=3}
ggplot(listings, aes(x=review_scores_rating, y=reviews_per_month)) + 
  geom_point() + 
  geom_smooth(method="lm")
```


### 2. Which neighborhoods have the highest median price?  Which neighborhoods have the highest variance or largest IQR?  Hint: Use `neighbourhood_cleansed` for neighborhoods.

**Answer:**  
Downtown has the highest median price.
```{r fig.width=6}
ggplot(listings,aes(x=reorder(neighbourhood_cleansed, price, FUN = median),
                    y=price)) +
  geom_boxplot() + coord_flip()
```

The Leather District has the highest variance and IQR
```{r fig.width=6}
#Variance
ggplot(listings,aes(x=reorder(neighbourhood_cleansed, price, FUN = var),
                    y=price)) +
  geom_boxplot() + coord_flip()+
  ggtitle("Variance ordered")

#IQR
ggplot(listings,aes(x=reorder(neighbourhood_cleansed, price, FUN = IQR),
                    y=price)) +
  geom_boxplot() + coord_flip()+
  ggtitle("IQR ordered")
```

### 3. Load the built-in dataset `mpg` with `data(mpg)`.  Use `?mpg` to learn more about the variables included.  Come up with a hypothesis or expectation you have about the data and check it with one of the methods used above. 

**Answer:**  
There are many possible solutions here!  Let's look at some basic information:

```{r}
data(mpg)
#?mpg
table(mpg$class)
table(mpg$year)
```

Hypothesis:
The proportion of compact cars increases over the two time periods.  Let's test it with a plot:

```{r fig.width=6}
ggplot(mpg,aes(x=as.factor(year),fill=class)) + geom_bar()
```

This might be a little clearer with some of our tools from week 3:

```{r fig.width=6}
mpg %>% filter(class == "compact") %>% 
  ggplot(aes(x=as.factor(year),fill=class)) + geom_bar()
```


### 4. Load the built-in dataset `msleep` with `data(msleep)`.  Use `?msleep` to learn more about the variables included.  Come up with a hypothesis or expectation you have about the data and check it with one of the methods used above.

**Answer:**  
There are many possible solutions here!  Let's look at some basic information:

```{r}
data(msleep)
#?msleep
summary(msleep$brainwt)
summary(msleep$sleep_rem)
```
 
Hypothesis:
Animals with larger brains experience more REM sleep.

```{r fig.width=6}
ggplot(data=msleep, aes(x=brainwt, y=sleep_rem))+
  geom_point()
```

There doesn't seem to be too much to this hypothesis!  What if we normalize both variables - considering instead the ratio of brain weight to body weight and REM sleep to total sleep?

```{r fig.width=6}
ggplot(data=msleep, aes(x=brainwt/bodywt, y=sleep_rem/sleep_total))+
  geom_point()+
  geom_smooth(method="lm")
```

It looks like there may be a slight downward trend here, but we don't have strong evidence.  Sometimes EDA doesn't lead us to strong conclusions!
