---
title: "Exercise Solutions: Models"
author:
- name: University of North Carolina at Chapel Hill
output: 
  html_document:
    theme: spacelab
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

```{r download, message=FALSE}
# Load the tidyverse package.
library(tidyverse) 
```

### 1. Explore the datasets included with the built-in `datasets` package by running `data(package="datasets")`.  You can get more information about any given dataset with `?<dataset name>`.  Choose a dataset with at least a few numeric variables.

**Answer:**  
```{r}
#data(package="datasets")
?swiss
```


### 2. Use `str`, `head`, `summary` and any of our EDA techniques to explore the dataset.  Pick a continuous outcome variable of interest and choose some predictors.

**Answer:**  
```{r}
str(swiss)
head(swiss)
summary(swiss)
```


```{r}
library(GGally)
ggpairs(swiss)
```

Let's use `Fertility` as our outcome variable and use the other 5 variables as predictors.

### 3. Fit an `lm` with your selected variables.

**Answer:**  
```{r}
swiss_lm <- lm(data=swiss, Fertility ~ Agriculture + Examination + Education + Catholic + Infant.Mortality)
#or we could use the . shortcut for all of the other variables:
#swiss_lm <- lm(data=swiss, Fertility~.)
summary(swiss_lm)
```


### 4. Use your intial `lm` fit to create a [stargazer table](https://unc-libraries-data.github.io/R-Open-Labs/Extras/Stargazer/Stargazer.html) output.  Embed in an R Markdown document or output html.  Use `?stargazer` to learn about and then change one or more default settings.

**Answer:**  
Remember to include `{r results="asis"}` as your header in R Markdown.
```{r results="asis"}
library(stargazer)
stargazer(swiss_lm, type="html", title="Swiss Fertility Model", style ="ajps")
```

