---
title: "Extra - Regression Tables with Stargazer"
output: 
  html_document:
    toc: true
    theme: spacelab
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Necessary Packages

```{r}
#install.packages("stargazer")
library(stargazer)
```


## Formula notation in R

```{r}
data(iris)
mod <- lm(data=iris,Sepal.Length~Petal.Length+Petal.Width+Species)
summary(mod)
```

`lm` above is R's function for fitting Ordinary Least Squares (OLS) regressions. There are many other specialized regression models available in R, but fortunately the vast majority use similar notation for formulas.

The arguments given to `lm` are:

* Formula: `y~x`
    + R uses `~` in place of the equals sign in formulas
    + R automatically includes an intercept term
    + This formula is therefore equivalent to: 
    
      $$y=intercept + slope * x$$
* Data: 
    + The dataframe containing the variables of interest.
    

|Symbol|Role|Example|Equivalent|
|----|-----------|-----------|-----------------|
|`+`|Add variable|`mpg~vs+disp`|$$mpg = intercept + \beta_1 vs + \beta_2 disp$$|
|`*`|Interactions|`mpg~vs*disp`|$$mpg = intercept + \beta_1 vs + \beta_2 disp + \beta_3 vs*disp$$|
|`.`|Include all variables in dataframe| `mpg~.` | $$mpg = intercept + \beta_1 cyl + \beta_2 disp + ... + \beta_{10} carb$$|
|`-`|Exclude variable|`mpg~.-disp-hp`|$$mpg = intercept + \beta_1 cyl + \beta_2 drat + ... + \beta_{8} carb$$|


 Let's fit a simple model with the `iris` built-in data:

## Formatting `lm` style model results - `stargazer` package

Once you've fit a linear or some other model, you may want to report results. The `stargazer` package makes this relatively simple to do, especially in an R Markdown document. The below code will produce a common model summary format for a journal or presentation.  

The code block has the R markdown option `{r results = "asis"}`, which instructs R Markdown to use the HTML code that `stargazer` produces as part of the output document.

```{r results = "asis", message=FALSE, warning=FALSE, results=FALSE} 
mod <- lm(data=iris,Sepal.Length~Species)
mod1 <- lm(data=iris,Sepal.Length~Petal.Width+Species)
mod2 <- lm(data=iris,Sepal.Length~Petal.Length+Petal.Width+Species)

stargazer(mod,mod1,mod2, type = "html",  #we use html output to match our planned R Markdown output
     title = "My iris models")
```

<center>
```{r results = "asis", message=FALSE, warning=FALSE, echo=FALSE} 
mod <- lm(data=iris,Sepal.Length~Species)
mod1 <- lm(data=iris,Sepal.Length~Petal.Width+Species)
mod2 <- lm(data=iris,Sepal.Length~Petal.Length+Petal.Width+Species)

stargazer(mod,mod1,mod2, type = "html",  #we use html output to match our planned R Markdown output
     title = "My iris models")
```
</center>

We can also write the table directly to a file with the `out` argument:

```{r  results = 'hide'}
stargazer(mod, type = "html", out = "regression.html" ,title = "My iris model")
```

## Resources

There's a useful cheatsheet for `stargazer` and its myriad customization options [here.](https://www.jakeruss.com/cheatsheets/stargazer/)

## Exercise

* Pick a sample dataset (call `data()` to get a list of available datasets).
* Pick some variables and use `lm` to fit an OLS model.
* Create a stargazer table output.  Embed in an R Markdown document or output html.  
* Use `?stargazer` to learn about and then change one or more default settings.

