---
title: "GIS in R"
author: 
  name: University of North Carolina at Chapel Hill
execute:
  echo: true
format: 
  html:
    theme: spacelab
    toc: true
    toc-location: left
    page-layout: article
---

## Background on R

This workshop assumes a basic familiarity with the R programming language and some exposure to GIS concepts like projections.

If you'd like a refresher on R, we suggest the [R For Data Science](https://r4ds.hadley.nz/) free e-book and the lessons from our [beginR workshops](https://tarheels.live/beginr/lessons/), including [setup instructions for R and R Studio](https://unc-libraries-data.github.io/R-Open-Labs/week1_Introduction/R_OpenLabs_Introduction.html#setup-r-r-studio).

## Important packages

### Basic packages

As a refresher, in R we install packages from CRAN (The Comprehensive R Archive Network) with the following code:

`install.packages('tidyverse')`

You'll only need to install packages once, then you'll load them into your session with:

```{r}
#| message: false
#| warning: false
library(tidyverse)
```

We'll use the `tidyverse` family of packages to do non-spatial data manipulation today.

#### A note on pipes:

Today we'll use the base R pipe (`|>`) as our default. Some attendees may be more familiar with the `magrittr` pipe (`%>%`). The two work very similarly and should usually be interchangeable. If you'd like to use the base R pipe with R Studio's pipe shortcut (CTRL/CMD + SHIFT + m), you can change the default setting in **Tools \> Global Options \> Code \> Use native pipe operator, \|\> (requires R 4.1+)**.

### Spatial packages

For spatial processing we'll use the following packages today:

```         
install.packages(c('sf','nngeo','tidycensus','tigris','arcgislayers','tmap','ggmap'))
```

```{r}
#| message: false
#| warning: false
#Spatial processing
library(sf)
library(nngeo)

#Getting spatial data
library(tidycensus)
library(tigris)
library(arcgislayers)

#Thematic mapping
library(tmap)
library(ggmap)
```

The `sf` packages will be our main spatial processing tool for today. We'll use `nngeo` to do some nearest-neighbor calculations.

We'll explore some of the most common packages for easily loading spatial data: `tidycensus` for census data, `tigris` for census boundaries, and \`arcgislayers for loading data from ESRI Data Services.

Finally we'll use `ggplot` from the tidyverse to do some basic mapping and explore extensions with `tmap` for thematic mapping and `ggmap` for baselayers.

## Importing Data:

### CSVs / Tabular data

Most people first encounter spatial data through spatial points inside a larger table. These can be CSV or any other common tabular format.

### Shapefiles and Geopackages

Shapefiles are an older format with some significant limitations (like variable name lengths!), but they're important to be familiar with because they're very very common and quirky. Despite the name, a shapefile is actually a *collection of files that have to be stored together* - anywhere from 3 to 12 files!  It's good practice to share them as a zipped folder.

Geopackages are a newer file format that avoid many of the limitations of shapefiles. We'll use geopackages for locally storing outputs from our R scripts.

There are lots of other spatial formats out there like KML, GeoJSON, PostGIS and many more!  R can work with many of these but we won't work with them all today.

#### A note on File Geodatabases

### Online Data:

#### `tidycensus` and `tigris`

The [`tidycensus`](https://walker-data.com/tidycensus/) and [`tigris`](https://github.com/walkerke/tigris) packages make it easy to pull census data directly from Census APIs into your R session. `tigris` pulls TIGER/Line data with boundaries and a few fields (FIPS code, Land and Water Area, etc.) for a variety of census geographies. This is great for various mapping and spatial processing tasks and gives more control over feature resolution.

Sometimes you'll also want to retrieve census *data* along with spatial information. `tidycensus` makes it easy to retrieve tables from a variety of census datasets, along with spatial data when relevant.

```{r}
ac5_vars <- load_variables(year=2024, dataset = "acs5")
```


```{r}
nc_counties <- get_acs(geography = "county", state = "NC", table = "B01001A",
        output = "wide", geometry = TRUE, year = 2024, cache_table = TRUE)
```

```{r}
plot(st_geometry(nc_counties))
```


```{r}
nc <- states() |> filter(STUSPS %in% c("NC","SC","VA","GA"))
```
```{r}
ggplot(data = nc, aes(fill = STUSPS!="NC"))+
  geom_sf(color='grey80')+
  theme_minimal()
```


```{r}
plot(st_geometry(nc))
```


#### ESRI Data Services with `arcgislayers`

```{r}
furl <- paste0(
  "https://sampleserver6.arcgisonline.com/arcgis/rest/services/",
  "EmergencyFacilities/FeatureServer/0"
)

# use field aliases as column names
sf_layer <- arc_read(furl, alias = "replace")

```

## Quick note on projections

### Checking and specifying projections

### Reprojecting data

## Data Operations

### Filtering

### Join on attributes

## Spatial Operations

### Spatial Filtering

### Buffering

### Spatial Joins

#### Nearest Neighbors with `nngeo`

## Plotting Data:

### `sf` and `ggplot`

### `tmap`

## Other useful packages:

### Rasters

There are two major raster packages to know about! `terra` is the successor to the `raster` package and works well for most simple rasters. `stars` is a better fit for cubes of data - think rasters with many many layers covering lots of different time points. `stars` itself is short for "Spatiotemporal Arrays"!

### Leaflet

### Other data formats

geoparquet, duckDB

### `spatialreg`
